% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J.~Seguinot et al.}
\date{February 15, 2018}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\referee}[2][]{\bigskip\noindent\textcolor{darkblue}{\textbf{#1:} #2}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Andreas Vieli,

Thank you very much for your feedback and support for our manuscript.

    \referee[p.~1, l.~6]{%
        to be correct it should probably say: '...resolution of 1 to 2 km'.}

    We rephrased to ``using horizontal resolutions of 2 and 1\,km and up to
    576~processors''.

    \referee[p.~2, l.~12]{%
        '...indicating SEVERAL tens of ...'. And change 800 000 years to 800 ka
        (and delete the bracket (800ka) on next line.}

    Done.

    \referee[p.~5, Fig.~1c caption]{%
        maybe say that this is at the 1km resolution used in the model.}

    Good idea, we've added this precision.

    \referee[p.~11, l.~6]{%
        '...by a east-west...'}

    Thanks, we've corrected.

    \referee[p.~15]{%
        I am just wondering and maybe this is to be considered in a later stage
        after a first round of reviews: in the section 4.2 i wonder if you
        should not briefly consider the discrepancy of the erratics from the
        southern Valais that reached the solothurner lobe... (work of Jouvet et
        al). it seems with your modelling you do not get Valais ice to flow to
        Solothurn (only aare valley)}

    Although not entirely obvious from the figures this is a correct
    observation. We've added a sentence about this.

    \referee[p.~18, Fig.~6 caption l.~4]{%
        should it not say 50m bands (instead of 500m bands)?}

    That is right, we've corrected.

    \referee[p.~18, l.~5--7]{%
        should you not mention the work by Becker here (his Phd work), as he
        also did preliminary sensitivity studies exploring a range of factors
        parameters (including sliding/rheology) for a part of the alpine ice
        sheet. Or am I wrong here?}

    We've added a sentence about that.

\end{document}
